A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-21 tex/codatadp.tex
Minor clarification

First, we need additional terminology concerning $\mu$-terms.
For a $\mu$-term $t$ with subterm $u$,
the \emph{interpretation of\/ $u$ in\/ $t$} is the $\mu$-term $t\interp{u}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \MU x.\; \const{C}( \vec u ) \in t$ indicates that $\MU x.\; \const{C}( \vec u )$ is the subterm of $t$ that binds this occurrence of variable $x$:
\[\begin{array}{r@{}c@{}l}
t\interp{x}^M & {} \,=\, {} &
\textstyle\begin{cases}
    \MU x.\; \const{C}( t\interp{ \vec u}^{M \mathrel\cup \{ x \}}) & \text{if }x \not\in M, x \mapsto \MU x.\; \const{C}( \vec u ) \in t  \\[-\jot]
    x & \text{otherwise}
\end{cases} \\
t\interp{\MU x.\; \const{C}( \vec u )}^M & = &
\textstyle\begin{cases}
    \MU x.\; \const{C}( t\interp{ \vec u }^{M \mathrel\cup \{ x \}} ) & \text{if }x \not\in M \\[-\jot]
    x & \text{otherwise}
\end{cases}
\end{array}\]
We write $t\interp{u}$ to abbreviate $t\interp{u}^\emptyset$.

@@

We now define a class of models that we consider for a saturated set of equalities~$\Ec$.
Let $\M( \tau )$ denote the interpretation type $\tau$ in $\M$,
that is, a set of domain elements for that type.
For a term $t$, let $\M( t )$ denote the interpretation of $t$ in $\M$,
that is, an element of $\M( \tau )$ where $\tau$ is the type of $t$.
In the following,
\emph{all} types are interpretated as a set of $\mu$-terms.
This is done to simplify the presentation.
Since all values of types in $\Data$ and $\Nondata$ are acyclic,
a typical representation for these values can be obtained by
dropping each of its $\mu$-bindings. 
For instance, the $\mu$-term $\mu x. \const{0}$ representing the integer constant zero
is simply $\const{0}$ after dropping the $\mu$-binding of $x$.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \Val^\ast \ec{ u_1 }, \ldots, \Val^\ast \ec{ u_n } )}$, call this term $t_u$.
By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\ast$, we have $\Val^\ast \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
where $\Val^\ast \ec{t}\interp{w_i} = \Val^\ast \ec{u_i}$ for $i = 1, \ldots, n$.
For each $i = 1, \ldots, n$, let $u_i'$ be the $i$\vvthinspace th argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $t_u\interp{u_i'} \vsim \Val^\ast \ec{u_i}$,
and thus $t_u\interp{u_i'} \vsim \Val^\ast \ec{t}\interp{ w_i }$. % for each $i = 1, \ldots, n$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\M( t ) \vsim t_u \vsim \Val^\ast \ec{t}$, and
we have $\M( t ) \vsim \Val^\ast \ec{t}$.

If $t$ is a selector term of the form $\const s^k_j( u )$,
since \rn{Split} does not apply,
$\ec{u}$ must contain a term of the form $\const C_{j'}( \const s^1_{j'}( u ), \ldots, \const s^{n}_{j'}( u ) )$ for some $j'$.
By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\ast$, we have $\Val^\ast \ec{u}$ is of the form
$\MU {\Varec{u}}.\; \const C_{j'}( w_1, \ldots, w_n )$,
where $\Val^\ast \ec{u} \interp{ w_i } = \Val^\ast \ec{ \const s^i_{j'}( u )}$ for $i = 1, \ldots, n$. 
If $j = j'$, then $\M( t )$ is $\alpha$-equivalent with $\Val^\ast \ec{u} \interp{w_k}$, which is equal to $\Val^\ast \ec{ \const s^k_j( u )}$,
which is $\Val^\ast \ec{t}$.
If $j \neq j'$, we are free to interpret $\const s^k_j( \M( u ) )$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\Val^\ast \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}.3,
pairs of wrongly applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

>>>>

First, we need to define the set of interpretations of the theory $\thD$,
which requires additional terminology concerning $\mu$-terms.
For a $\mu$-term $t$ with subterm $u$,
the \emph{interpretation of\/ $u$ in\/ $t$} is the $\mu$-term $t\interp{u}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \MU x.\; \const{C}( \vec u ) \in t$ indicates that $\MU x.\; \const{C}( \vec u )$ is the subterm of $t$ that binds this occurrence of variable $x$:
\[\begin{array}{r@{}c@{}l}
t\interp{x}^M & {} \,=\, {} &
\textstyle\begin{cases}
    \MU x.\; \const{C}( t\interp{ \vec u}^{M \mathrel\cup \{ x \}}) & \text{if }x \not\in M, x \mapsto \MU x.\; \const{C}( \vec u ) \in t  \\[-\jot]
    x & \text{otherwise}
\end{cases} \\
t\interp{\MU x.\; \const{C}( \vec u )}^M & = &
\textstyle\begin{cases}
    \MU x.\; \const{C}( t\interp{ \vec u }^{M \mathrel\cup \{ x \}} ) & \text{if }x \not\in M \\[-\jot]
    x & \text{otherwise}
\end{cases}
\end{array}\]
We write $t\interp{u}$ to abbreviate $t\interp{u}^\emptyset$.

@@

We now define the class of interpretations for $\thD$.
In the following, $\M( \tau )$ denotes the interpretation type $\tau$ in $\M$,
that is, a set of domain elements for that type.
For a term $t$, $\M( t )$ denotes the interpretation of $t$ in $\M$,
that is, an element of $\M( \tau )$ where $\tau$ is the type of $t$.
In the following,
\emph{all} types are interpretated as a set of $\mu$-terms.
This is done to simplify the presentation.
Since all values of types in $\Data$ and $\Nondata$ are acyclic,
a typical representation for these values can be obtained by
dropping each of its $\mu$-bindings. 
For instance, the $\mu$-term $\mu x. \const{0}$ representing the integer constant zero
is simply $\const{0}$ after dropping the $\mu$-binding of $x$.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \Val^\ast \ec{ u_1 }, \ldots, \Val^\ast \ec{ u_n } )}$, call this term $t_u$.
%By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\ast$, we have $\Val^\ast \ec{t}$ is a term
%of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
%where $\Val^\ast \ec{t}\interp{w_i} = \Val^\ast \ec{u_i}$ for $i = 1, \ldots, n$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by our construction of $\Val^\ast$ we have $\Val^\ast \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$ 
where for each $i = 1, \ldots, n$, 
$\muvar(w_i) = {\Varec{u_i}}$ and thus by Lemma~\ref{lem:model-completion}.2, 
$\Val^\ast \ec{t}\interp{w_i} = \Val^\ast \ec{u_i}$.
For each $i = 1, \ldots, n$, let $u_i'$ be the $i$\vvthinspace th argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $t_u\interp{u_i'} \vsim \Val^\ast \ec{u_i}$.
Thus $t_u\interp{u_i'} \vsim \Val^\ast \ec{t}\interp{ w_i }$. % for each $i = 1, \ldots, n$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\M( t ) \vsim t_u \vsim \Val^\ast \ec{t}$, and
we have $\M( t ) \vsim \Val^\ast \ec{t}$.

If $t$ is a selector term of the form $\const s^k_j( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\const C_{j'}( \const s^1_{j'}( u ), \ldots, \const s^{n}_{j'}( u ) )$ for some $j'$.
%By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\ast$, we have $\Val^\ast \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{j'}( w_1, \ldots, w_n )$,
%where $\Val^\ast \ec{u} \interp{ w_i } = \Val^\ast \ec{ \const s^i_{j'}( u )}$ for $i = 1, \ldots, n$. 
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by our construction of $\Val^\ast$ we have $\Val^\ast \ec{u}$ is of the form
$\MU {\Varec{u}}.\; \const C_{j'}( w_1, \ldots, w_n )$,
where for each $i = 1, \ldots, n$,
$\muvar(w_i) = {\Varec{\const s^i_{j'}( u )}}$ and thus by Lemma~\ref{lem:model-completion}.2, 
$\Val^\ast \ec{u} \interp{ w_i } = \Val^\ast \ec{ \const s^i_{j'}( u )}$. 
If $j = j'$, then $\M( t )$ is $\alpha$-equivalent with $\Val^\ast \ec{u} \interp{w_k}$, which is equal to $\Val^\ast \ec{ \const s^k_j( u )}$,
which is $\Val^\ast \ec{t}$.
If $j \neq j'$, we are free to interpret $\const s^k_j( \M( u ) )$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\Val^\ast \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}.3,
pairs of wrongly applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.
